---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

options(tibble.print_min = 6, tibble.print_max = 6)
```

# comperes: Manage Competition Results

[![Build Status](https://travis-ci.org/echasnovski/comperes.svg?branch=master)](https://travis-ci.org/echasnovski/comperes) [![codecov](https://codecov.io/gh/echasnovski/comperes/branch/master/graph/badge.svg)](https://codecov.io/gh/echasnovski/comperes)

`comperes` offers a pipe (`%>%`) friendly set of tools for storing and managing competition results. Understanding of __competition__ is quite general: it is a set of __games__ (abstract event) in which __players__ (abstract entity) gain some abstract __scores__ (typically numeric). The most natural example is sport results, however not the only one. For example, product rating can be considered as a competition between products as "players". Here a "game" is a customer that reviews a set of products by rating them with numerical "score" (stars, points, etc.).

This package leverages [dplyr](http://dplyr.tidyverse.org)'s grammar of data manipulation. Only basic knowledge is enough to use `comperes`.

## Overview

`comperes` provides the following functionality:

- __Store and convert__ competition results:
    - In _long format_ as a [tibble](http://tibble.tidyverse.org) with one row per game-player pair. Functions: `as_longcr()`, `is_longcr()`.
    - In _wide format_ as a `tibble` with one row per game with fixed amount of players. Functions: `as_widecr()`, `is_widecr()`.
- __Summarise__:
    - Compute _item summaries_ with functions using `dplyr`'s grammar. Functions: `summarise_item()`, `summarise_game()`, `summarise_player()`.
    - Compute and _join_ item summaries to data for easy transformation. Functions: `join_item_summary()`, `join_game_summary()`, `join_player_summary()`.
    - Use _common item summary functions_ with [rlang](https://rlang.r-lib.org/)'s [unquoting](http://rlang.r-lib.org/reference/quasiquotation.html) mechanism. Example: `. %>% summarise_player(!!!summary_funs["mean_score"])`.
- __Compute Head-to-Head values__ (a summary statistic of direct confrontation between two players) with functions also using `dplyr`'s grammar:
    - Store output in _long format_ as a `tibble` with one row per pair of players. Function: `h2h_long()`.
    - Store output in _matrix format_ as a matrix with rows and columns describing players and entries - Head-to-Head values. Function: `h2h_mat()`.
    - Use _common Head-to-Head functions_ with rlang's unquoting mechanism. Example: `. %>% h2h_mat(!!!h2h_funs["num_wins"])`.

## Installation

You can install `comperes` from CRAN with:

```{r cran-installation, eval = FALSE}
install.packages("comperes")
```

To install the most recent development version from GitHub use:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("echasnovski/comperes")
```

## Examples

### Store and Convert

We will be using `ncaa2005`, data from `comperes` package. It is an example competition results (hereafter - results) of an isolated group of Atlantic Coast Conference teams provided in book ["Who's #1"](https://www.amazon.com/Whos-1-Science-Rating-Ranking/dp/069116231X) by Langville and Meyer. It looks like this:

```{r ncaa2005, message = FALSE, warning = FALSE}
library(comperes)
ncaa2005
```

This is an object of class `longcr` which describes results in long form (_each row represents the score of particular player in particular game_). Because in this competition a game always consists from two players, more natural way to look at `ncaa2005` is in wide format:

```{r ncaa2005-wide}
as_widecr(ncaa2005)
```

This converted `ncaa2005` into an object of `widecr` class which describes results in wide format (_each row represents scores of all players in particular game_). All `comperes` functions expect either a data frame with results structured in long format or one of supported classes: `longcr`, `widecr`.

### Summarise

With `compere` the following summaries are possible:

```{r item-summary, message = FALSE, warning = FALSE}
ncaa2005 %>%
  summarise_player(min_score = min(score), mean_score = mean(score))

# Using list of common summary functions
library(rlang)
ncaa2005 %>%
  summarise_game(!!!summary_funs[c("sum_score", "num_players")])
```

Supplied list of common summary functions has `r length(summary_funs)` entries, which are quoted expressions to be used in `dplyr` grammar:

```{r summary_funs, message = FALSE, warning = FALSE}
summary_funs

ncaa2005 %>% summarise_player(!!!summary_funs)
```

To modify scores based on the rest of results one can use `join_*_summary()` functions:

```{r join-item-summary, message = FALSE, warning = FALSE}
suppressPackageStartupMessages(library(dplyr))
ncaa2005_mod <- ncaa2005 %>%
  join_player_summary(player_mean_score = mean(score)) %>%
  join_game_summary(game_mean_score = mean(score)) %>%
  mutate(score = player_mean_score - game_mean_score)

ncaa2005_mod

ncaa2005_mod %>% summarise_player(mean_score = mean(score))
```

This code modifies `score` to be average player score minus average game score. Negative values indicate poor game performance.

### Head-to-Head

Computation of Head-to-Head performance is done with `h2h_long()` (output is a tibble; allows multiple Head-to-Head values per pair of players) or `h2h_mat()` (output is a matrix; only one value per pair of players).

Head-to-Head functions should be supplied in `dplyr` grammar but for players' matchups: direct confrontation between __ordered__ pairs of players (including playing with themselves) stored in wide format:

```{r matchups, message = FALSE, warning = FALSE}
ncaa2005 %>% get_matchups()
```

Typical Head-to-Head computation is done like this:

```{r h2h, message = FALSE, warning = FALSE}
ncaa2005 %>%
  h2h_long(
    mean_score_diff = mean(score1 - score2),
    num_wins = sum(score1 > score2)
  )

ncaa2005 %>% h2h_mat(mean(score1 - score2))
```

Supplied list of common Head-to-Head functions has `r length(h2h_funs)` entries, which are also quoted expressions:

```{r h2h_funs, message = FALSE, warning = FALSE}
h2h_funs

ncaa2005 %>% h2h_long(!!!h2h_funs)
```

To compute Head-to-Head for only subset of players or include values for players that are not in the results, use factor `player` column:

```{r h2h-factor, message = FALSE, warning = FALSE}
ncaa2005 %>%
  mutate(player = factor(player, levels = c("Duke", "Miami", "Extra"))) %>%
  h2h_mat(!!!h2h_funs["num_wins"], fill = 0)
```

